---
title: "make_QC_plots"
output: html_document
date: "2025-11-06"
---

Plots to make --

RNA:
- UMIs per cell (neurons and glia)
- Genes per cell (neurons and glia)
- Percent mitochondrial (neurons and glia)
- (Maybe doublet rates?) (cell curation)
- UMIs per Class/Group/Cluster

ATAC:
- UMIs per cell (neurons and glia)
- FRiP, TSS enrichment, nucleosomal ratio *[PYTHON]*
- ATAC fragment size distributions (per library, lineplot mean with quantiles shaded)
- ATAC metagene at TSSes (if I have time, but that's my thing so)

# Env

```{r}
# probably not using most of these
suppressPackageStartupMessages({
    library(dplyr)
    library(ComplexHeatmap)
    library(circlize)
    library(RColorBrewer)
    library(ggplot2)
    library(patchwork)
    library(ggrepel)
    library(ggpubr)
    library(viridisLite)
})


theme_md_classic <- function() {
    theme_classic(
        base_size = 6, 
        base_family = "ArialMT",
        base_line_size = 0.25,
        base_rect_size = 0.25
    ) %+replace% theme(
        plot.title  = element_text(hjust = 0.5, face = "bold", size = 8),
        axis.title  = element_text(color = "black", size = 6),
        axis.text   = element_text(color = "black", size = 6),
        axis.ticks  = element_line(color = "black"),
        legend.text = element_text(size = 6),
        strip.text  = element_text(size = 8),
        strip.background = element_blank(),
        complete = TRUE
    )
}
```

## Paths

```{r}
pdir <- "/Volumes/krienen/marm_hmba_integration/250602_reprocess_and_recluster"
path_cellr_summary <- file.path(pdir, "metadata/cellranger-arc_summary.csv")
path_cur_stats     <- file.path(pdir, "metadata/cell_curation_stats.csv")
path_qc_table      <- file.path(pdir, "qc/library_qc_pass_fail_table.csv")
path_rna_cmeta     <- file.path(pdir, "data/rna/rna_clean_0715_cell_metadata.csv")
path_atac_cmeta    <- file.path(pdir, "data/atac/atac_cell_meta.csv")
path_anno          <- file.path(pdir, "metadata/preprint_freeze/subcortex_anno_table_preprint.csv")
```

```{r}
outdir <- file.path(pdir, "qc/plots")
if (!dir.exists(outdir))
    dir.create(outdir)

if (!dir.exists(outdir, "curated"))
    dir.create(file.path(outdir, "curated"))

if (!dir.exists(outdir, "cellranger"))
    dir.create(file.path(outdir, "cellranger"))
```

## Imports

```{r}
cellr_summary <- read.csv(path_cellr_summary)
cur_stats     <- read.csv(path_cur_stats)
qc_table      <- read.csv(path_qc_table)
rna_cmeta     <- read.csv(path_rna_cmeta, row.names = "X")
atac_cmeta    <- read.csv(path_atac_cmeta, row.names = "index")
anno          <- read.csv(path_anno, row.names = "marm_cluster_id")
```

fragment size distribution data from multiqc:

```{r}
atac_ins_sizes <- read.delim(file.path(pdir, "qc/multiqc_data/mqc_atac_insert_size_plot.txt"))

# import like tuples (x,y value pairs with just numbered columns)
atac_ins_sizes <- atac_ins_sizes %>%
    tidyr::pivot_longer(-Sample, names_to = "col", values_to = "pair") %>%
    # Remove parentheses and split into two numeric columns
    mutate(
        pair = gsub("[()]", "", pair),
        insert_size = as.numeric(sub(",.*", "", pair)),
        count = as.numeric(sub(".*,", "", pair))
    ) %>%
    select(Sample, insert_size, count)
```


```{r}
### NOT USED
# Also, ATAC reads in consensus peaks by cluster:
# atac_reads_peaks <- read.csv(
#     file.path(pdir, "qc/data/atac_reads_in_consensus_peaks_by_cluster.csv")
# )
```

## Drop remaining Junk

Most is already gone so this is appropriate. The QC will basically show what's
curated in the library (the cellranger summary etc. will take care of earlier
steps)

```{r}
# > nrow(rna_cmeta)
# [1] 681571
```

That's 681571 cells. Manuscript says 678917 cells.

```{r}
# > all(rna_cmeta$Cluster_v4 %in% rownames(anno))
# [1] TRUE
# > sum(anno[rna_cmeta$Cluster_v4, "Subcortex_Class_v4"] != "Junk")
# [1] 678917

### that's it. remaining Junk can be seen in anno table
```

```{r}
# > nrow(anno)
# [1] 1811
anno <- subset(anno, Subcortex_Class_v4 != "Junk")
# > nrow(anno)
# [1] 1737
```

```{r}
rna_cmeta <- rna_cmeta[rna_cmeta$Cluster_v4 %in% rownames(anno), ]
# > nrow(rna_cmeta)
# [1] 678917
```

ATAC is a subset anyway:

```{r}
# > nrow(atac_cmeta)
# [1] 651082
atac_cmeta <- atac_cmeta[rownames(atac_cmeta) %in% rownames(rna_cmeta), ]
# > nrow(atac_cmeta)
# [1] 632920
```


## Join anno table into cell metadata

### RNA

Makes things a little easier.

```{r}
# > colnames(rna_cmeta)
#  [1] "barcoded_cell_sample_name"                 "tissue_name"                              
#  [3] "dissociated_cell_sample_name"              "donor_name"                               
#  [5] "mit_name"                                  "sample_name"                              
#  [7] "donor_slab_tile"                           "hemisphere"                               
#  [9] "alignment_job_id"                          "nCount_RNA"                               
# [11] "nFeature_RNA"                              "percent.mt"                               
# [13] "percent.Rpl"                               "percent.Hbb"                              
# [15] "scDblFinder.class"                         "scDblFinder.score"                        
# [17] "scDblFinder.weighted"                      "scDblFinder.cxds_score"                   
# [19] "S.Score"                                   "G2M.Score"                                
# [21] "Phase"                                     "old_barcode"                              
# [23] "ABCmouse_MapMyCells_CLAS_label"            "ABCmouse_MapMyCells_SUBC_label"           
# [25] "ABCmouse_MapMyCells_CLUS_label"            "Siletti_MapMyCells_Supercluster_label"    
# [27] "Siletti_MapMyCells_Cluster_label"          "Siletti.reannot_MapMyCells_Class_label"   
# [29] "Siletti.reannot_MapMyCells_Subclass_label" "Siletti.reannot_MapMyCells_Cluster_label" 
# [31] "Cluster_v4"                                "Cluster_v3"                               
# [33] "Subcortex_Class_v4"                        "Subcortex_Group_v4"    

# > colnames(anno)
#  [1] "Subcortex_Class_v4"                  "Subcortex_Group_v4"                  "Cluster_v4_size"                    
#  [4] "Cluster_NT"                          "Figure_Panel"                        "Panel_sections"                     
#  [7] "Panel_Purpose"                       "Subcortex_color_hex_cluster"         "Subcortex_color_hex_cluster_display"
# [10] "preprint_analysis_include"           "Subcortex_Class_v4_BG"               "Cluster_Dominant_Region"            
# [13] "Group_Dominant_Region"               "Group_NT" 
```

```{r}
# > names(rna_cmeta)[names(rna_cmeta) %in% names(anno)]
# [1] "Subcortex_Class_v4" "Subcortex_Group_v4"

rna_cmeta[names(rna_cmeta) %in% names(anno)] <- NULL
```

```{r}
rna_cmeta <- cbind(rna_cmeta, anno[rna_cmeta$Cluster_v4, ])
```

### ATAC

```{r}
# > names(atac_cmeta)[names(atac_cmeta) %in% names(anno)]
# [1] "Subcortex_Class_v4" "Subcortex_Group_v4"

atac_cmeta[names(atac_cmeta) %in% names(anno)] <- NULL
```

```{r}
atac_cmeta <- cbind(atac_cmeta, anno[atac_cmeta$Cluster_v4, ])
```


## Add a neuron/glia column

```{r}
rna_cmeta <- mutate(
    rna_cmeta,
    neur_glia = ifelse(
        Cluster_NT == "",
        "glia",
        "neurons"
    )
)
```

```{r}
atac_cmeta <- mutate(
    atac_cmeta,
    neur_glia = ifelse(
        Cluster_NT == "",
        "glia",
        "neurons"
    )
)
```

# RNA [curated]

## Curation stats

### Columns

I did this before...

```{r}
cur_stats
```

```{r}
subset(cur_stats, sample_name %in% rna_cmeta$sample_name) %>% 
    .[, names(.) != "sample_name"] %>% 
    lapply(sum) %>% 
    reshape2::melt() %>% 
    setNames(c("cells", "step")) %>% 
    mutate(step = factor(step, levels=step)) %>% 
    ggplot(
        aes(step, cells/1e6)
    ) + 
    # geom_line() + 
    geom_col(color="black", fill="gray75", linewidth=0.25) + 
    labs(x=NULL, y="cells (millions)") +
    scale_y_continuous(limits=c(0,1.25), expand=c(0,0)) +
    theme_md_classic() + 
    theme(
        axis.text.x = element_text(angle=30, hjust=1)
    )
```

```{r}
ggsave(
    file.path(outdir, "rna_curation_colplot.pdf"),
    width=60, height=50, units="mm"
)
```

### Boxplot

```{r}
subset(cur_stats, sample_name %in% rna_cmeta$sample_name) %>% 
    setNames(., sub("cells_under_1e.05_umi", "cells_under_100k_umi", names(.))) %>% 
    reshape2::melt(id.var = "sample_name", value.name = "cells") %>% 
    ggplot(
        aes(variable, cells)
    ) + 
    geom_violin(linewidth=0.25) +
    geom_boxplot(linewidth=0.25, outliers=FALSE, width=0.3) + 
    labs(x=NULL, y="cells per library", title = "Multiome RNA Cell Curation") +
    # scale_y_continuous(limits=c(0,1.25), expand=c(0,0)) +
    theme_md_classic() + 
    theme(
        axis.text.x = element_text(angle=30, hjust=1)
    ) -> p0
p0
```

```{r}
ggsave(
    file.path(outdir, "rna_curation_boxplot.pdf"),
    width=60, height=50, units="mm"
)
```

## UMIs per cell

```{r}
p1 <- ggplot(
    rna_cmeta,
    aes(neur_glia, log10(nCount_RNA))
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "UMIs per cell (log10)",
        title = "Multiome RNA"
    )
p1
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_UMIs_per_cell.pdf"),
    p1,
    width=30,
    height=35,
    units="mm"
)
```

## Genes per cell

```{r}
p2 <- ggplot(
    rna_cmeta,
    aes(neur_glia, log10(nFeature_RNA))
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "genes per cell (log10)",
        title = "Multiome RNA"
    )
p2
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_genes_per_cell.pdf"),
    p2,
    width=30,
    height=35,
    units="mm"
)
```

## Percent mitochondrial

```{r}
p3 <- ggplot(
    rna_cmeta,
    aes(neur_glia, percent.mt/100)
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    scale_y_continuous(
        labels = scales::percent
    ) + 
    labs(
        x = NULL,
        y = "% mitochondrial",
        title = "Multiome RNA"
    )
p3
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_percent_mito.pdf"),
    p3,
    width=30,
    height=35,
    units="mm"
)
```

## Cells per Group, Cluster

```{r}
table(rna_cmeta$neur_glia, rna_cmeta$Subcortex_Group_v4) %>% 
    as.data.frame %>% 
    subset(Freq>0) %>% 
    setNames(c("neur_glia", "group", "cells")) %>% 
    ggplot(
        aes(neur_glia, log10(cells))
    ) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "cells per Group (log10)",
        title = "Multiome RNA"
    ) -> p4
p4
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_cells_per_Group.pdf"),
    p4,
    width=30,
    height=35,
    units="mm"
)
```

```{r}
table(rna_cmeta$neur_glia, rna_cmeta$Cluster_v4) %>% 
    as.data.frame %>% 
    subset(Freq>0) %>% 
    setNames(c("neur_glia", "group", "cells")) %>% 
    ggplot(
        aes(neur_glia, log10(cells))
    ) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "cells per Cluster (log10)",
        title = "Multiome RNA"
    ) -> p5
p5
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_cells_per_Cluster.pdf"),
    p5,
    width=30,
    height=35,
    units="mm"
)
```

## UMIs per Group, Cluster

```{r}
p6 <- ggplot(
    aggregate(rna_cmeta, nCount_RNA ~ neur_glia + Subcortex_Group_v4, FUN=sum) ,
    aes(neur_glia, log10(nCount_RNA))
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "UMIs per Group (log10)",
        title = "Multiome RNA"
    )
p6
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_UMIs_per_Group.pdf"),
    p6,
    width=30,
    height=35,
    units="mm"
)
```


```{r}
p7 <- ggplot(
    aggregate(rna_cmeta, nCount_RNA ~ neur_glia + Cluster_v4, FUN=sum) ,
    aes(neur_glia, log10(nCount_RNA))
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "UMIs per Cluster (log10)",
        title = "Multiome RNA"
    )
p7
```

```{r}
ggsave(
    file.path(outdir, "curated/RNA_UMIs_per_Cluster.pdf"),
    p7,
    width=30,
    height=35,
    units="mm"
)
```

# ATAC [curated]

## UMIs per cell

```{r}
p8 <- ggplot(
    atac_cmeta,
    aes(neur_glia, log10(n_fragment))
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "UMIs per cell (log10)",
        title = "Multiome ATAC"
    )
p8
```

```{r}
ggsave(
    file.path(outdir, "curated/ATAC_UMIs_per_cell.pdf"),
    p8,
    width=30,
    height=35,
    units="mm"
)
```

## (Skip): Peaks per cell

I don't have this, would have to load the peak matrix. Not sure anyone asks
for this anyway.

## TSS Enrichment (SnapATAC2)

```{r}
p9 <- ggplot(
    atac_cmeta,
    aes(neur_glia, tsse)
) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(width=0.3, outliers=FALSE, linewidth=0.25) + 
    theme_md_classic() + 
    labs(
        x = NULL,
        y = "TSS enrichment (per cell) (SnapATAC2)",
        title = "Multiome ATAC"
    )
p9
```

```{r}
ggsave(
    file.path(outdir, "curated/ATAC_Snap_TSSE_per_cell.pdf"),
    p9,
    width=30,
    height=35,
    units="mm"
)
```

## (skip) FRiP (cluster-level)

### Calculate

I think the fragments in the cell metadata is actually reads, but the "reads in
peaks" is actually insertion events, so divide by two

```{r}
# atac_reads_peaks$total_reads <- aggregate(n_fragment ~ Cluster_v4, atac_cmeta, FUN=sum) %>%
#     `rownames<-`(., .[[1]]) %>% 
#     .[atac_reads_peaks$marm_cluster_id, "n_fragment"]

# atac_reads_peaks %>% 
#     mutate(
#         frip = 0.5*reads_in_peaks / total_reads
#     )
```

Maybe skip this, use cellranger-arc peaks (valid QC).

# ATAC [cellranger]

## Fractions frags in cells, TSS, peaks

```{r}
cellr_summary %>% 
    subset(Sample.ID %in% atac_cmeta$sample_name) %>% 
    .[, c(
        "Sample.ID", 
        "ATAC.Fraction.of.high.quality.fragments.in.cells",
        "ATAC.Fraction.of.high.quality.fragments.overlapping.TSS",
        "ATAC.Fraction.of.high.quality.fragments.overlapping.peaks"
    )] %>% 
    setNames(c(
        "sample_name",
        "frac. fragments in cells",
        "frac. fragments on TSS",
        "frac. fragments in peaks"
    )) %>% 
    reshape2::melt() %>% 
    ggplot(
        aes(variable, value)
    ) + 
    geom_violin(linewidth=0.25) + 
    geom_boxplot(linewidth=0.25, width=0.3) + 
    coord_flip() +
    scale_y_continuous(limits = c(0,1), expand=c(0,0)) +
    labs(
        x = NULL,
        y = NULL,
        title = "ATAC (cellranger-arc)"
    ) +
    theme_md_classic()
```


```{r}
ggsave(
    file.path(outdir, "cellranger/ATAC_fractions_frags_in_peaks_tss_cells.pdf"),
    width=50,
    height=35,
    units="mm"
)
```


# Fragment size distributions

```{r}
atac_ins_sizes %>% 
    subset(Sample %in% atac_cmeta$sample_name) %>% 
    split(., .$Sample) %>% 
    lapply(mutate, count = count / max(count)) %>% 
    unname %>% 
    do.call(rbind, .) %>% 
    ggplot(
        aes(insert_size, count, group=Sample)
    ) + 
    geom_line(alpha=0.05, linewidth=0.25) + 
    labs(
        y = "scaled frequency",
        x = "insert size (bp)",
        title = "ATAC Fragment Sizes"
    ) +
    theme_md_classic()
```

```{r}
ggsave(
    file.path(outdir, "ATAC_insert_dist.pdf"),
    width=40,
    height=35,
    units="mm"
)
```

